---
title: "COMP4033 Assignment1-group7"
output:
  html_document: default
  pdf_document: default
date: "2026-03-26"
---

```{r setup, include=FALSE}
library(tidyverse)
oral_data <- read.csv("oral_cancer_prediction_dataset.csv")
```

## 1. Dataset Structure

This section provides an overview of the dataset structure, including the number of variables and their data types.

```{r}
str(oral_data)
```

## 2. Variables in Dataset

The dataset contains the following variables:

```{r}
names(oral_data)
```

## 3. Preview of Data

Below are the first 15 rows of the dataset:

```{r}
head(oral_data, 15)
```

# 4. User-defined function

Calculate a risk score based on the person's data with tobacco, alcohol, HPV, family history, and compromised immune system.

```{r}
risk_score <- function(tobacco, alcohol, hpv, family_history, immune_system) {
  (tobacco == "Yes") * 2 +
  (alcohol == "Yes") +
  (hpv == "Yes") * 2 + 
  (family_history == "Yes") + 
  (immune_system == "Yes")
}

risk_score(oral_data$Tobacco.Use[1],
           oral_data$Alcohol.Consumption[1],
           oral_data$HPV.Infection[1],
           oral_data$Family.History.of.Cancer[1],
           oral_data$Compromised.Immune.System[1])
```

# 5. Filter rows based on critera

Get rows with stage 1 cancer and tumor sizes less than 5 cm.

```{r}
output = oral_data %>% filter(
       oral_data$Cancer.Stage == "1",
       oral_data$Tumor.Size..cm.<"5")
head(output, 15)
```

# 6. Identify dependent & independent variables

```{r}
# Independent var: Tumor Size -- Dependent var: Cost of Treatment
head(data.frame(oral_data$Tumor.Size..cm., oral_data$Cost.of.Treatment..USD.), 15)

# Independent var: Diet (fruits, veg intake) -- Dependent var: Cancer Stage
head(data.frame(oral_data$Diet..Fruits...Vegetables.Intake., oral_data$Cancer.Stage), 15)
```

# 7. Data cleaning and adjustments

Remove missing values in your dataset. Identify and remove duplicated data in your dataset.

Reorder multiple rows in descending order. Rename some of the column names in your dataset.

```{r}
clean_data = oral_data %>% filter(!is.na(Cost.of.Treatment..USD.))
clean_data = clean_data %>% distinct()

clean_data = clean_data %>% arrange(desc(Cost.of.Treatment..USD.))

clean_data = clean_data %>% rename(
  Drinker = Alcohol.Consumption,
  Treatment.Cost.USD = Cost.of.Treatment..USD.
)
```

# 8. Add new variables to the data frame

Add a "Cost per cm" column based on Treatment Cost divided by Tumor Size.

```{r}
clean_data = clean_data %>% mutate(cost.per.cm = Treatment.Cost.USD / Tumor.Size..cm.)
```

# 9. Create a training set

Create a training set using random number generator engine with seed `1234` with 5% the size of the main data set.

```{r}
set.seed(1234) 
training_data <- clean_data %>% sample_frac(0.05, replace = FALSE)
```

# 10. Summary of the main data set

```{r}
summary(clean_data)
```

# 11. Mean and range of treatment costs

```{r}
mean(clean_data$Treatment.Cost.USD, na.rm = TRUE)
range(clean_data$Treatment.Cost.USD, na.rm = TRUE)
```

# 12. Scatter plot

Scatter plot of Tumor Size and Treatment Cost.

```{r}
ggplot(data = training_data, aes(x = Tumor.Size..cm., y = Treatment.Cost.USD)) + 
  geom_point(color = "steelblue", size = 1.2)
```

# 13. Bar plot

Bar plot of displaying number of individuals at each cancer stage.

```{r}
ggplot(data = training_data, aes(x = Cancer.Stage, fill = Gender)) + geom_bar() + 
  ylab("Number of Individuals")
```

# 14. Pearson correlation

A correlation near 1 implies high correlation; the dataset has about ~0.75 correlation, so there is a strong linear relationship between Treatment Cost and Tumor Size.

```{r}
cor(clean_data$Treatment.Cost.USD, clean_data$Tumor.Size..cm., method="pearson")
```
