---
title: "Assignment 1 – Data Analysis using R Programming"
output:
  html_document: default
  pdf_document: default
date: "2026-03-28"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
oral_data <- read.csv("oral_cancer_prediction_dataset.csv")
```

**Group 7**: Esther Agyemang, Omar Alamodi, Aliha Bashir, Hetal Bulsara,
Andy Huang, James Huynh, Nasteha Mohamed, and Tina Shen

George Brown Polytechnic\
COMP 4033 – Health Informatics Data Analytics

## Related Links

- **GitHub repository:**
    <https://github.com/CabbageCanFly/comp4033-2026-03-a1>\
- **RPubs link:** <https://rpubs.com/cabbagecanfly/1413996>\
- **Dataset source:**
    <https://www.kaggle.com/datasets/ankushpanday2/oral-cancer-prediction-dataset>

## 1. Dataset Structure

This dataset contains patient-level information related to oral cancer cases.

The `str()` function shows that the dataset contains 84,922 observations and 
25 variables. It includes a mix of numeric and character variables representing 
demographics, risk factors, clinical symptoms, and treatment outcomes.

```{r}
str(oral_data)
```

\newpage

## 2. Variables in Dataset

The `names()` function lists all 25 variables in the dataset. These include 
demographic information (e.g., Age, Gender, Country), risk factors (e.g., 
Tobacco Use, Alcohol Consumption, HPV Infection), clinical indicators, and 
outcome variables such as survival rate and treatment cost.

```{r}
names(oral_data)
```

## 3. Preview of Data

The `head()` function displays the first 15 rows of the dataset, providing a 
quick overview of how the data is organized.

```{r}
head(oral_data, 15)
```

## 4. User-Defined Function

A user-defined function is created to calculate a simple risk score based on 
selected risk factors, including tobacco use, alcohol consumption, HPV 
infection, family history, and immune system status. Each factor contributes a 
weighted value to produce an overall score for demonstration purposes.

```{r}
risk_score <- function(tobacco, alcohol, hpv, family_history, immune_system) {
  (tobacco == "Yes") * 2 +
  (alcohol == "Yes") +
  (hpv == "Yes") * 2 + 
  (family_history == "Yes") + 
  (immune_system == "Yes")
}

risk_score(oral_data$Tobacco.Use[1],
           oral_data$Alcohol.Consumption[1],
           oral_data$HPV.Infection[1],
           oral_data$Family.History.of.Cancer[1],
           oral_data$Compromised.Immune.System[1])
```

## 5. Filter Rows Based on Criteria

An example filter is applied based on the following conditions:

- Stage 1 cancer  
- Tumor sizes less than 5 cm  
- From Sri Lanka  
- Over the age of 60  
- Female  
- No tobacco use  
- No betel quid use

The narrow criteria result in a small subset of patients, in this case older 
females from Sri Lanka with early-stage cancer and generally fewer high-risk 
behaviors.

```{r}
oral_data %>% filter(
  oral_data$Cancer.Stage == 1,
  oral_data$Tumor.Size..cm. < 5,
  oral_data$Country == "Sri Lanka",
  oral_data$Age > 60,
  oral_data$Gender == "Female",
  oral_data$Tobacco.Use == "No",
  oral_data$Betel.Quid.Use == "No")
```

\newpage

## 6. Identifying Dependent & Independent Variables

The dataset contains multiple potential independent and dependent variables; a 
subset has been selected for demonstration purposes.

Independent variable: **Tumor Size**  
Dependent variable: **Cost of Treatment**

```{r}
head(data.frame(oral_data$Tumor.Size..cm., oral_data$Cost.of.Treatment..USD.), 15)
```

Independent variable: **Diet (Fruits, Vegetables Intake)**  
Dependent variable: **Cancer Stage**

This relationship is explored for demonstration purposes, recognizing that it may be indirect.

```{r}
head(data.frame(oral_data$Diet..Fruits...Vegetables.Intake., oral_data$Cancer.Stage), 15)
```

\newpage

## 7. Data Cleaning and Adjustments

Basic data cleaning steps are demonstrated, including **removing missing values**, 
**eliminating duplicate** records, **sorting** the data in descending order by 
treatment cost, and **renaming** selected columns for clarity. In this dataset, no 
missing values or duplicate records were observed. Columns such as alcohol 
consumption and treatment cost were renamed for readability as part of the 
demonstration.

```{r}
clean_data = oral_data %>% filter(!is.na(Cost.of.Treatment..USD.))
clean_data = clean_data %>% distinct()

clean_data = clean_data %>% arrange(desc(Cost.of.Treatment..USD.))

clean_data = clean_data %>% rename(
  Drinker = Alcohol.Consumption,
  Treatment.Cost.USD = Cost.of.Treatment..USD.
)
head(clean_data, 5)
```

## 8. Adding New Variables to the Data Frame

A new variable, **cost per cm**, is created by dividing **treatment cost** by 
**tumor size**. This demonstrates how derived variables can be added to the 
dataset for further analysis.

```{r}
clean_data = clean_data %>% mutate(cost.per.cm = Treatment.Cost.USD / Tumor.Size..cm.)
head(clean_data, 5)
```

## 9. Creating a Training Set

A training set is created by randomly sampling 5% of the dataset using a fixed 
seed (1234) to ensure reproducibility. This demonstrates how subsets of data 
can be generated for modeling or testing purposes.

```{r}
set.seed(1234) 
training_data = clean_data %>% sample_frac(0.05, replace = FALSE)
head(training_data, 5)
```

\newpage

## 10. Summary of the Main Data Set

The `summary()` function provides an overview of the dataset.

From the output, many values for tumor size and treatment cost are zero, which 
may indicate placeholder or missing entries. The median tumor size and cancer
stage are also zero, suggesting many cases are recorded at early or minimal 
levels. Additionally, a large number of missing values appear in 
cost per cm, due to division by zero when tumor size is zero.

```{r}
summary(clean_data)
```

## 11. Statistical Functions on the Dataset

**Mean**, **median**, and **range** of treatment costs.

Basic statistical measures such as mean, median, and range are calculated for 
treatment cost. The mean cost is approximately 39,110 USD, while the median is 
0, indicating that a large portion of the dataset contains zero-cost entries. 
The range spans from 0 to 159,988 USD, showing a wide variation in treatment 
costs.

```{r}
mean(clean_data$Treatment.Cost.USD, na.rm = TRUE)
median(clean_data$Treatment.Cost.USD, na.rm = TRUE)
range(clean_data$Treatment.Cost.USD, na.rm = TRUE)
```

R does not have a built-in `mode()` function for statistical mode (most
frequent value). The `mode()` function in R refers to the data type of an 
object. So the following code was executed to determine the statistical **mode**
of cancer stage in the dataset; this code can be put in a user-defined function 
for reusability, but the following is included to illustrate the steps taken.

The most common cancer stage in the dataset is 0, which aligns with earlier 
observations that many cases are recorded at early or minimal stages.

1. `table()` is used to count how many times each cancer stage appears.

```{r}
cancer_stage_count = table(clean_data$Cancer.Stage)
cancer_stage_count
```

2. Sort the counts from highest to lowest.

```{r}
cancer_stage_count = sort(cancer_stage_count, decreasing = TRUE)
cancer_stage_count
```

3. **Obtain the cancer stage with the highest count.**

```{r}
names(cancer_stage_count)[1]
```


\newpage

## 12. Scatter Plot

A scatter plot is created to visualize the relationship between 
**survival rate (5 years)** and **treatment cost**.

The plot shows a slight negative trend, where higher survival rates are 
generally associated with lower treatment costs. However, the relationship is 
weak, as there is significant variability in costs across all survival levels.

```{r}
ggplot(data = training_data, aes(x = Survival.Rate..5.Year...., y = Treatment.Cost.USD)) + 
  geom_point(color = "steelblue", size = 1.2)
```

\newpage

## 13. Bar Plot

A bar plot is created to display the **number of individuals at each cancer stage**, 
grouped by **gender**.

From the plot, stage 0 has the highest number of individuals, with counts 
decreasing as cancer stage increases. Males appear to have higher counts than 
females across all stages, although the overall distribution trend is similar 
for both groups.

```{r}
ggplot(data = clean_data, aes(x = Cancer.Stage, fill = Gender)) + geom_bar() + 
  ylab("Number of Individuals")
```

## 14. Pearson Correlation

A correlation value close to 1 or -1 indicates a strong linear relationship. 
The Pearson correlation between **treatment cost** and **survival rate** is 
approximately -0.8, indicating a strong negative correlation, where higher 
survival rates are generally associated with lower treatment costs. While the 
scatter plot of the same variables (Section 12) shows high variability, the 
overall downward trend still results in a strong negative correlation.

```{r}
cor(clean_data$Treatment.Cost.USD, clean_data$Survival.Rate..5.Year...., method="pearson")
```

<div style="height: 200px;"></div>
